---
title: "Postpartum Community Remodeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Postpartum Community Remodeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

### Data and Problem Context

```{r}
library(fs)
library(tidyverse)
library(fido2)
library(mdsine)
library(mbtransfer)
library(patchwork)
library(glue)
th <- theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_rect(fill = NA, color = "#0c0c0c", linewidth = 0.6),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.position = "bottom"
  )
theme_set(th)
set.seed(20230524)
```

```{r}
tr_fun <- function(method, hyper = list()) {
  function(x) {
    if (method == "mdsine") {
      mdsine(ts, hyper$taxonomy, n_clusters=hyper$n_clusters)
    } else if (method == "mbtransfer") {
      mbtransfer(ts, P = hyper$P, Q = hyper$Q)
    } else if (method == "fido") {
      fido(ts, sigma = hyper$sigma, rho = hyper$rho, design = hyper$design)
    }
  }
}

evaluate_splits <- function(ts, f, ...) {
  fits <- list()
  ts_preds <- list()

  # in-sample performance
  fits[["in-sample"]] <- f(ts)
  ts_missing <- subset_values(ts, 1:12)
  ts_preds[["in-sample"]] <- predict(fits[["in-sample"]], ts_missing, ...)
  subject_data(ts_preds[["in-sample"]]) <- subject_data(ts_missing) |>
    mutate(subject = as.factor(subject))

  # out-of-sample performance
  fits[["out-of-sample"]] <- f(ts[1:25])
  ts_preds[["out-of-sample"]] <- predict(fits[["out-of-sample"]], ts_missing[26:49], ...)
  subject_data(ts_preds[["out-of-sample"]]) <- subject_data(ts[26:49]) |>
    mutate(subject = as.factor(subject))
  list(fits = fits, ts_preds = ts_preds)
}

plot_predictions <- function(ts, ts_preds) {
  #browser()
  reshaped_preds <- map_dfr(ts_preds, ~ reshape_preds(ts, .), .id = "generalization") |>
   filter(h > 0, h < 50) |>
    mutate(h = 10 * floor(h / 10))

  p <- reshaped_preds |>
    mutate(h = glue("lag {h}")) |>
    ggplot() +
    geom_abline(slope = 1, col = "#400610") +
    geom_point(aes(y, y_hat, col = generalization), size = .7, alpha = .6) +
    facet_grid(factor(quantile, rev(levels(quantile))) ~ h, scales = "free_y") +
    labs(x = expression(y), y = expression(hat(y)), col = "Generalization") +
    scale_x_continuous(expand = c(0, 0), n.breaks = 3) +
    scale_y_continuous(expand = c(0, 0), n.breaks = 3) +
    scale_color_manual(values = c("#A65883", "#6593A6")) +
    guides("color" = guide_legend(override.aes = list(size = 4, alpha = 1))) +
    theme(
      axis.text = element_text(size = 10),
      panel.spacing = unit(0, "line"),
      strip.text.y = element_text(angle = 0, size = 12),
      strip.text.x = element_text(angle = 0, size = 12),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 11),
    )

  correlation <- reshaped_preds |>
    group_by(h, quantile, generalization) |>
    summarise(correlation = round(cor(y, y_hat), 4)) |>
    pivot_wider(names_from = "generalization", values_from = "correlation") |>
    arrange(desc(quantile), h)

  list(p = p, correlation = correlation)
}
```

To construct as `ts_inter` object, we combine the following datasets:

* `reads`: A `data.frame` of taxonomic compositions. Rows are samples and columns
are taxa.
* `samples`: A `data.frame` of sample-level descriptors. This is expected to have
`sample` and `time` columns. `sample` must match the rownames across `reads` and
`interventions`.
* `subject_data`: A `data.frame` of subject-level descriptors. This can be used
to store information that is not time varying.
* `interventions`: A `data.frame` of whose  rows are samples and columns are
interventions. This can be either a binary matrix of whether a given
intervention (column) was present in a sample. An input series can
also be continuously valued -- see the aquaculture vignette ofr an example.

We created these datasets by lightly processing the raw data from the published
repository. Our processing script is available here. Note that we converted all
the subject-level data to numeric variables -- `mbtransfer` does not know how to
handle factor inputs, so they must be coded first.

```{r}
reads <- read_csv("https://figshare.com/ndownloader/files/40322782/reads.csv") |>
  column_to_rownames("sample")
samples <- read_csv("https://figshare.com/ndownloader/files/40322776/samples.csv")
subject_data <- read_csv("https://figshare.com/ndownloader/files/40322773/subject.csv")
interventions <- read_csv("https://figshare.com/ndownloader/files/40322770/interventions.csv") |>
  column_to_rownames("sample")
```

````{r}
ts <- reads |>
  ts_from_dfs(interventions, samples, subject_data) |>
  interpolate(delta = 14, method = "linear")

ts
```

```{r}
subject_data(ts) <- subject_data(ts) |>
  select(subject, BirthControlYesNo) |>
  mutate(BirthControlYesNo = ifelse(BirthControlYesNo == "Yes", 1, 0))
```

```{r}
hyper <- list(sigma = 1, rho = 1, design = "~ BirthControlYesNo + time + subject")
predict <- fido2::predict
#method_eval <- evaluate_splits(ts, tr_fun("fido", hyper), design = hyper$design)
#summaries <- plot_predictions(ts, method_eval$ts_preds)
#save(method_eval, summaries, file = "postpartum_fido.rda")
```

```{r}
load("case_studies/postpartum_fido.rda")
ggsave("postpartum_fido_predictions.png", width = 10, height = 6)
write_csv(summaries$correlation, file = "postpartum_fido_correlations.csv")
```

We can now do the same comparison with MDSINE2.

```{r}
# a hack -- mdsine requires a sample at time 0
for (i in seq_along(ts)) {
  min_above <- min(which(ts[[i]]@time >= 0))
  ts[[i]]@time[min_above] <- 0
}

taxonomy <- data.frame(tax_id = 1:ncol(reads), name = paste0("tax", 1:ncol(reads)), kingdom = "bacteria", phylum = NA, class = NA, order = NA, family = NA, genus = NA, species = NA, sequence = NA)
hyper <- list(taxonomy = taxonomy, n_clusters = as.integer(15))
predict <- mdsine::predict
#method_eval <- evaluate_splits(ts, tr_fun("mdsine", hyper))
load("postpartum_mdsine.rda")
#summaries <- plot_predictions(ts, method_eval$ts_preds)
#save(method_eval, summaries, file = "postpartum_mdsine.rda")
```