---
title: "Hospitalome Preliminary Study"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(tidyverse)
library(fs)
theme_set(theme_bw())
data_dir <- path("../microbiome_transfer_data/hospitalome/")
```

## Figure 1

Let's first try to reproduce Figure 1 from the paper.

```{r}
antibiotics <- read_csv(data_dir / "tbldrug.csv")
abundances <- read_csv(data_dir / "tblcounts_asv_melt.csv")
samples <- read_csv(data_dir / "tblASVsamples.csv")
temperature <- read_csv(data_dir / "tbltemperature.csv")
taxonomy <- read_csv(data_dir / "tblASVtaxonomy_silva132_v4v5_filter.csv")
```

This figure is just about patient 1511, so let's filter.

```{r}
patient <- 1511
antibiotics <- antibiotics %>%
  filter(PatientID == patient)
temperature <- temperature %>%
  filter(PatientID == patient)
samples <- samples %>%
  filter(PatientID == patient)

abundances <- abundances %>%
  right_join(samples)
```

Here is their antibiotics figure.

```{r}
ggplot(antibiotics) +
  geom_segment(aes(StartDayRelativeToNearestHCT, reorder(Factor, StartDayRelativeToNearestHCT, min), xend = StopDayRelativeToNearestHCT, yend = Factor), size = 2, col = "#535353") +
  coord_cartesian(xlim = c(-5, 20))
```

Here is their temperature figure.

```{r, fig.width = 6, fig.height = 2}
ggplot(temperature, aes(DayRelativeToNearestHCT, MaxTemperature)) +
  geom_line() +
  geom_point(aes(col = MaxTemperature > 101)) +
  scale_color_manual(values = c("black", "red"), guide = FALSE) +
  xlim(-5, 20) +
  ylim(98, 102)
```

Now let's make the microbiome composition figure. It's note quite the same,
since we're just aggregating to Genus relative abundances (not a mix of genera
and phyla). But we can still see the disappearance of Ruminococcus and blooming
of Lactococcus.

```{r}
genera <- abundances %>%
  left_join(select(taxonomy, ASV, Genus)) %>%
  group_by(SampleID, Genus) %>%
  summarise(Count = sum(Count)) %>%
  group_by(SampleID) %>%
  mutate(
    Total = sum(Count),
    proportion = Count / Total
  ) %>%
  left_join(select(samples, SampleID, DayRelativeToNearestHCT))

top_genera <- genera %>%
  group_by(Genus) %>%
  summarise(total = sum(Count)) %>%
  arrange(-total) %>%
  pull(Genus) %>%
  head(10)

genera <- genera %>%
  mutate(Genus_group = factor(Genus, levels = top_genera))
```

```{r}
ggplot(genera) +
  geom_col(aes(DayRelativeToNearestHCT, proportion, fill = Genus_group, col = Genus_group)) +
  xlim(-5, 20)
```

